Re-Allocation of Resources during Releases \titlenote{Copyright note}

### 2013-12-01 paper1/paper1.tex
major literary correction made the paper, new research question added. rq done 1-5.

\renewcommand{\labelenumi}{Q\theenumi:}
\begin{enumerate}
\item What is the release process used by the project? \newline
We will be giving a qualitative description of Linux Kernel development in brief to give anser to this question.
\item How many developers are allocated troughout different segments of a release period? What significant difference can be observed between the segments of an entire release period? \newline
Answer to this question will give us a big picture of Linux Kernel development community as well as the contrubution of developers in different periods within a stable release period. We will calculate Jaccard Distance to differentiate the behavior of developers during RTR and normal development period.
\item Do developers work on different areas of the system around the time of release? \newline
We want to see if developers are looking at different types of files in the code-base around the time of release which is being considered as the merge period or merge window\cite{linux_kernel}. We are expecting this would be the RTR are in a release time where technical debt may be introduced.
\item Do developers work in others' code in a high proportion during the rush period?
Many developers are working on many files. We will find out which files has been churned mostly by which developer so that we can understand how developers are distributed among the system.
\item Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)? \newline
To answer this question we will look into the frequency of file churns made into files. Difference or distance of the files between normal development period and the RTR period or merge period of a release.
\end{enumerate}

@@

\subsection{Time-Span of Releases}
Another information that we require is what are the durations of the releases of Linux kernel development. To achieve that we joined the table ``git\_refs\_tags'' where all the releases are stored including the splitted release numbers and the table ``git\_commit'' to give us all the dates of commits for each and every releases, so that we can easily find out the duration between two consecutive releases (including release candidates RC). This information will significantly help us because we need to see what is the general development period and what are the merge periods which can be treated as RTR period within a release period. We also need to understand how developers are working, what is the impact of their changes made in the code-base during a particular release period later on. Table 1 shows a small portion of this information. Preior to calculate the developers' areas in secion 4.4 we require this information.

@@

\subsubsection{Developers' Area in a Merge Period}
Merge Perid can be determined from the previous release date (of any of major, minor or micro releases) to the date of rc-1 (i.e.\ date of the first candidate release). For example if the date of release for ``linuxv2.6.12''  is 2005-06-17 and date of pushing the rc-1 for the next release ``linuxv2.6.13'' is 2005-06-29 then this time period of of 12 days is going to be being called the merge window. We are considering this segment as a Rush To Release or RTR because thousands of commits gets pushed in the main branch during this time and developers remain in a rush to pack up merging all the branches those are to go with this release. Table 2 shows merge periods of some releases.

@@

\subsubsection{Developers' Areas in Release Period}
As like merge period we can extract developers working in code files during release period. Release Period here in Linux Kernel development process is being considered as since the date of rc-1 release (i.e.\ first release candidate opening the gate for the development of the next release) to the date of the next stable release. For example if the date of rc-1 for ``linuxv2.6.13'' after releasing ``linuxv2.6.12'' is 2005-06-29 and the date of publishing the release ``linuxv2.6.13'' is 2005-08-28 then we are calling this 60 days of development period for the release as the Release Period. Table 4 is showing some development periods of different releases.

@@

\subsection{Finding Code Ownership}
So many developers are working from different development communities throught different development branches. We tried to investigate the "code ownership" \cite{mockus_case_study} evolved in Linux Kernel development. Here code ownership is going to be represented as a percentage of making churns or commits by a developer with respect to the total churns or commits made to a particular file. From the data that we have for complete releases (``linuxv2.6.13'' to ``linuxv3.11'') we found 75138 distinct files that have been churned or commited by developers. We investigate the total number of commits, total number of churns for each and every file, we also investigate how many developers contributed for these commits and changes. This information yet does not tell about the ownership, we need to find out ownership for each developer in different development areas in a release cycle. We already have the collection of developers working in merge period and release period within a release cycle and stored them with corresponding files they have worked on, how many commits a developer made to a particular file and how many changes he/she has made. We now update those information with the ownership information. Equation 1 is the base for calculating ownership.
\begin{equation}\omega=\frac{n}{N}*100\end{equation}
Where $\omega$ is the ownership and n is the number of total changes made by an atuthor of a file, N is the total number of changes made for a particular file.
To determine a developer be a owner of a file we are considering that if the developer makes changes more than 80\% of the total churn that has been made by all contributors so far for that file, then that developer can be called as a owner of that file, similarly as C. Gutwin did \cite{gutwin_awareness} for finding out the main developers' group for a project. We already mentioned that 14569 individual files didn't have any change (addition or deletion) but was being committed at least once. So we found them in our release wise datasets that's why in addition to Gutwin here we are involving commits to determine ownerships but only for those files that do not have churn information. If a file has no addition or deletion information in it's commit log but has been committed by some one at least once then that or those authors may have the ownership for that file. So, we have made a little change in equation 1 which becomes like:
\begin{equation} \omega = \left\{ \begin{array}{l l} n/N & \quad \text{if $N > 0$}\\ c/C & \quad \text{if $N = 0$} \end{array} * 100 \right.\end{equation}
Where c is the number of commits made by the author and C is the total number of commits made for the particular file.
We calculate code ownership for both Merge Period and Release Period. We also canculate general ownership for all the developers for all the files they have worked ever. Table 6 represents a part of the data which shows the ownerships of the developers in each release cycle.

@@

In the early 1990's the Linux kernel development was not so busy affair \cite{linux_kernel}. There were very small number of users and developers involved at that time. Developers use a loosely time-based release process. Linux publishes it's major kernel release in every two or three months. If we look at figure 2 the upper Line represents how the stable releases are going on since 2005. From the figure we can see the average duration of each stable release is 62.47 although ``linuxv2.6.24'' was significantly hight as exception, it took 92.8 days. The bottom line which is much steady than the upper one is representing the durations of merge period in each stable release. If we put the releases in a time domain we can see a picture shown in figure 3 where we see that the stable releases are all going on maintaining almost a similar frequency and similar pattern. The highest picks are the merge periods then the smaller picks between the highest ones are representing the rc releases. For more clear view we have plotted only 100 releases of all 340 releases.

@@

In order to see how many developers contribute to the new featcher developments and how many contribute in merging and fixing, review and accepting codes from the community during the merge period, we observed the developers' area in Merge Period and developers' area in Release Period. Out of all 381152 commits belonging to all 340 releases 292926 commits were made during the time of merge period which is 76.85\% (more than 2/3) of the total. These commits were made by 14621 individual developers. We now keeping only the records belong to release 2.6.13 to 3.11 because we have complete information for thse releases, so we had to remove the commit records belong to linuxv2.6.12 and linuxv3.12 and later. 11209 developers work in merge period and 7625 developers work around the release period. We found that 4908 developers contribute in both merge period and the release period that means near about half (43.78\%) of the developers work during the merge period are shared that means they work in both merge and release periods.

@@

We are supposed to find the answers for the two research questions together in this section. Developers work for the software continuously. Some are working for the core development some people are for the fixing and minor enhancements and some for feature developments, planning to merge new features with any up coming release but we would like to understand what number of files they change during the time of regular development period and how many during the time of RTR. How many of them they really own and how many files they work that they do not own. We also want to know about the code ownership of the developers around the time of release and general code ownership i.e.\ which files a developer has worked making more than 80\% chruns in lifetime.

Since the release of ``linuxv2.6.13'' till ``linuxv3.11'' 2889 developers out of 14621 worked for the files they owned. Although they have worked for other files that they do not own. So the percentage of working with owned files is 19.82\% through out the life time (release 2.6.13 to 3.11). We examine the contribution in writing code and making churns by the developers to understand the behaviour and to see if the impact of this 19.82\% on the graph. Figure 7 plots the cumulative representation of churnes files and percentage of owned files per developer for a particular release ``linuxv2.6.13''. If I say in details, in the increase of every 10 developers it is observed that developers work increasingly more unknown files than the files they own or they are very familier to. We see the similar picture for other releases too.
So we can say that in Linux Kernel development developers work in more unfamilier files than the most familier (i.e.\ more than 80\%) ones.
\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3.4in]{cumulativeChurnOwnershipPercent.png}
\caption{\small \sl Cumulative Number of Churns vs Percentage of Owned Files in release 2.3.13}
\end{center}
\end{figure}

To understand the distribution of the developers around the system we also found that during these release periods 8057 developers have experience of working in the merge meriod and 14621 developers worked in the release development period as we already mentioned earlier but we noted that 4884 developers work in both the parts. So almost half of the developers that work in merge period also work in the release development period. Figure 8 shows the Venn diagram.
\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3in]{devDistMPRDP.png}
\caption{\small \sl Developers' Distribution in Merge Period and Development Period}
\end{center}
\end{figure}

If we compare the percentage of working in owned files ($\omega$\textit{P}) in merge period and the percentage of working in owned files in release period ($\Omega$\textit{P}) for the life time for each developer works in the merge period then we see a significant difference as shown in figure 9.
\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3in]{ownedFilesMPRDPBox.png}
\caption{\small \sl Percentage of working in owned files during Merge Period and the that of life time}
\end{center}
\end{figure}

Now, to predict if developers work in others' code in a high proportion during the rush period we would like to observe the correlation of the \% of working in owned files in merge period with other independent variables ``churn percentage in merge period (\c{c}\textit{P})'', ``churn percentage in general (\c{C}\textit{P})'', ``average ownership ($\omega$\textit{Avg})'' and \% of wirking in owned files in general ($\Omega$\textit{P}). Table 6 shows the Spearman rank correlation among the \% of working with owned files and other variables. We can see there is no statistically significant positive relationship between the measures and the \% of working in owned files

All three correlations Kendall, Pearson and Spearman calculates that the percentage of working with owned files doesn't to any other variables strongly as we predicted seeing figure 7 having a very poor percentage of contribution (19.82\%) on owned files during the merge period of time. But we can see a significant strong relation between the \% of churn in general and \% of churn in merge period which is 0.7199 according to Kendall, 0.7028 according to Pearson and 0.8446 according to Spearman. Also the correlation between the \% of churn in general and the average ownership is strong according to Spearman model which is 0.7982. We are giving the priority to Spearman model because Spearman correlation is a more robust correlation technique \cite{nagappan_defect}.
If we consider the following models with our variables:
\renewcommand{\labelenumi}{M\theenumi:}
\begin{enumerate}
\item{Predictors: ($\omega$\textit{P}), $\Omega$\textit{P}}
\item{Predictors: ($\omega$\textit{P}), $\Omega$\textit{P}, $\omega$\textit{Avg}}
\item{Predictors: ($\omega$\textit{P}), $\Omega$\textit{P}, $\omega$\textit{Avg}, \c{c}\textit{P}}
\item{Predictors: ($\omega$\textit{P}), $\Omega$\textit{P}, $\omega$\textit{Avg}, \c{c}\textit{P}, \c{C}\textit{P}}
\end{enumerate}
then we see the linear regression model for the \% of working with owned code measure using all the predictors has an R-squared value 0.1132 with residual standard error is 32.52 on 8053 degrees of freedom, Adjusted R-squared value is 0.1139. F-statistic is 342.8 and the p-value is < 2.2e-16. R-square is a measure of variance in the dependent variable that is accounted for by the model built using the predictors \cite{nagappan_defect}. R-square is a measure of the fit for the given data set. It cannot be interpreted as the quality of the dataset to make future predictions.

\begin{table}[ht]
\caption{Cross Correlations}  % title of Table
\centering 						% used for centering table
\begin{tabular}{c c c c c c}			% centered columns (6 columns)
\hline\hline						%inserts double horizontal lines
% table heading
Measures  & $\omega$\textit{P} & $\Omega$\textit{P} & $\omega$\textit{Avg} & \c{c}\textit{P} & \c{C}\textit{P} \\ [0.5ex]
\hline 							% single horizontal line
% inserting body of the table
$\omega$\textit{P}		& 1.0000	& 0.2897	& 0.4496	& 0.3325	& 0.2826 \\
$\Omega$\textit{P}	& --			& 1.0000	& 0.7110	& 0.3979	& 0.5075 \\
$\omega$\textit{Avg}	& --			& --			& 1.0000 	& 0.6754 	& \textbf{0.7982} \\
\c{c}\textit{P}			& -- 			& -- 		 	& --			& 1.0000	& \textbf{0.8446} \\
\c{C}\textit{P}			& -- 			& -- 			& -- 			& -			& 1.0000 \\
[1ex]							% adds vertical space
\hline 							% inserts single line
\end{tabular}
\label{table:nonlin} 				% is used to refer this table in the text
\end{table}

@@

>>>>

\renewcommand{\labelenumi}{Q\theenumi:}
\begin{enumerate}
\item What is the release process used by the project? \newline
We will be giving a qualitative description of Linux Kernel development in brief to give anser to this question.
\item How many developers are allocated troughout different segments of a release period? What significant difference can be observed between the segments of an entire release period? \newline
Answer to this question will give us a big picture of Linux Kernel development community as well as the contrubution of developers in different periods within a stable release period. We will calculate Jaccard Distance to differentiate the behavior of developers during release period and development period.
\item Do developers work on different areas of the system around the time of release? \newline
We want to see if developers are looking at different types of files in the code-base around the time of release which is being considered as the release period or rc period. We are expecting this would be the RTR in a release cycle which may introduce technical debts.
\item Do developers work in others' code in a high proportion during the rush period?
Developers work on various files during release period. We will try to find out what is the percentage of dealing with files that developers own or they are most familier to. This will give us a sense of the nature of work during release time.
\item How long a commit take to get pushed into the main kernel? Does it differ in different periods of release?
In distributed development environment developers work in different branches they commit their codes to the branches they are associated to. Developers in that case do not directly push their changes to the main branch. We would like to see how much time it takes for a commit to the local branch to get pushed to the main branch. We also should understand how long it takes for regular development commits and how long it takes for the commits made around the release period.
\item Are there certain areas of the system that receive increased attention (i.e.\ do developers focus on a smaller set of files around releases)? \newline
To answer this question we will look into the frequency of file churns made into files. Difference or distance of the files between normal development period and the release period of a release.
\end{enumerate}

@@

\subsection{Time-Span of Releases}
Another information that we require is what are the durations of the releases of Linux kernel development. To achieve that we joined the table ``git\_refs\_tags'' where all the releases are stored including the splitted release numbers and the table ``git\_commit'' to give us all the dates of commits for each and every releases, so that we can easily find out the duration between two consecutive releases (including release candidates RC). This information will significantly help us because we need to see what is the development period and what are the release periods within a release cycle. We also need to understand how developers are working, what is the impact of their changes made in the code-base during a particular release period later on. Table 1 shows a small portion of this information. Preior to calculate the developers' areas in secion 4.4 we require this information.

@@

\subsubsection{Developers' Area in a Merge Period}
Merge Perid can be determined from the previous release date (of any of major, minor or micro releases) to the date of rc-1 (i.e.\ date of the first candidate release). For example if the date of release for ``linuxv2.6.12''  is 2005-06-17 and date of pushing the rc-1 for the next release ``linuxv2.6.13'' is 2005-06-29 then this time period of of 12 days is going to be being called the merge window. Thousands of commits get pushed in the main branch during this period. Table 2 shows merge periods of some releases.

@@

\subsubsection{Developers' Areas in Release Period}
As like merge period we can extract developers working in code files during release period. Release Period here in Linux Kernel development process is being considered as since the date of rc1 release (i.e.\ first release candidate opening the gate for the development of the next release) to the date of the next stable release. For example if the date of rc1 for ``linuxv2.6.13'' after releasing ``linuxv2.6.12'' is 2005-06-29 and the date of publishing the release ``linuxv2.6.13'' is 2005-08-28 then we are calling this 60 days of development period for the release as the Release Period. Table 4 is showing some development periods of different releases.

@@

\subsection{Finding Code Ownership}
So many developers are working from different development communities throught different development branches. We tried to investigate the "code ownership" \cite{mockus_case_study} evolved in Linux Kernel development. Here code ownership is going to be represented as a percentage of making churns by a developer with respect to the total churns made to a particular file. From the data that we have for complete releases (``linuxv2.6.13'' to ``linuxv3.11'') we found 70210 distinct files that have been churned by developers. We removed those files where no developer made any churn at all during these 39 releases. We investigate the total number of churns for each and every file, we also investigate how many developers contributed for these commits and changes. This information yet does not tell about the ownership, we need to find out ownership for each developer in different development areas in a release cycle. We already have the collection of developers working in merge period and release period within a release cycle and stored them with corresponding files they have worked on, how many commits a developer made to a particular file and how many changes he/she has made. We now update those information with the ownership information. Equation 1 is the base for calculating ownership.
\begin{equation}\omega=\frac{n}{N}*100\end{equation}
Where $\omega$ is the ownership and n is the number of total changes made by an atuthor of a file, N is the total number of changes made for a particular file.
To determine a developer be a owner of a file we are considering that if the developer makes changes more than 80\% of the total churn that has been made by all contributors so far for that file, then that developer can be called as a owner of that file, similarly as C. Gutwin did \cite{gutwin_awareness} for finding out the main developers' group for a project. We calculate code ownership for both Merge Period and Release Period. We also canculate general ownership for all the developers for all the files they have worked ever. Table 6 represents a part of the data which shows the ownerships of the developers in each release cycle.

@@

In the early 1990's the Linux kernel development was not so busy affair \cite{linux_kernel}. There were very small number of users and developers involved at that time. Developers use a loosely time-based release process. Linux publishes it's major kernel release in every two or three months. If we look at figure 2 the upper Line represents how the stable releases are going on since 2005. From the figure we can see the average duration of each stable release is 62.47 although ``linuxv2.6.24'' was significantly hight as exception, it took 92.8 days. The bottom line which is much steady than the upper one is representing the durations of merge period in each stable release. If we put the releases in a time domain we can see a picture shown in figure 3 where we see that the stable releases are all going on maintaining almost a similar frequency and similar pattern. The highest picks are the merge periods then the smaller picks between the highest ones are representing the rc releases. For more clear view we have plotted only 100 releases of all 340 releases including rc candidates.

@@

In order to see how many developers contribute to the new featcher developments and how many contribute in merging and fixing, review and accepting codes from the community during the merge period, we observed the developers' area in Merge Period and developers' area in Release Period. Out of all 381152 commits belonging to all 340 releases 292926 commits were made during the time of merge period which is 76.85\% (more than 2/3) of the total. These commits were made by 14621 individual developers. We now keeping only the records belong to release 2.6.13 to 3.11 because we have complete information for thse releases, so we had to remove the commit records belong to linuxv2.6.12 and linuxv3.12 and later. 11209 developers work in merge period and 7625 developers work around the release period.

@@

We are supposed to find the answers for the two research questions together in this section. Developers work for the software continuously. Some are working for the core development some people are for the fixing and minor enhancements and some for feature developments, planning to merge new features with any up coming release but we would like to understand what number of files they change during the time of development and merging period and how many during the release period. How many of them they really own and how many files they work that they do not own. We also want to know about the code ownership of the developers around the time of release as well as around the time of main development works that get merged during merge meriod.

In order to understand the distribution of the developers around the system we also found that during these release periods 11209 developers have experience of working in the merge meriod and 7625 developers worked in the release period as we already mentioned earlier but we noted that 4908 developers contribute in both merge period and the release period that means near about half (43.78\%) of the developers work during the merge period are shared that means they work for core developments and also bug fixing or minor enhancement works. Figure 7 representing the venn diagram.
\begin{figure}
\begin{center}
\includegraphics[height=1.3in,width=2.3in]{devDistMPRPvenn.png}
\caption{\small \sl Developers' Distribution in Merge Period and Development Period}
\end{center}
\end{figure}

From release ``linuxv2.6.13'' to ``linuxv3.11'' We examine the contribution in writing code by the developers to understand the behaviour during the two vital segments in a release cycle. Figure 8 plots the cumulative representation of the percentage of working with owned files ($\omega$\textit{MP}) during the merge period and the percentage of working with owned files ($\omega$\textit{RP}) during release period of ``linuxv2.6.13''. From the figure It can be observed that with the increase of every 10 files that developers working in, $\omega$\textit{RP} is going down apart from the line of $\omega$\textit{MP}. Up to a certain number may be around 2200 files developers mostly work with owned files. We see the similar picture for other releases too.
\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3.4in]{cumulFileOwnP.png}
\caption{\small \sl Cumulative percentage of working with owned files in MP and RP of release 2.3.13}
\end{center}
\end{figure}

If we simply compare the percentage of working in owned files during merge period ($\omega$\textit{MP}) and the percentage of working in owned files during release period ($\omega$\textit{RP}) for each developer then we see a significant difference of the mean lines as shown in figure 9. In release period developers work with less number of files that they own.

So we can say that in Linux Kernel development developers work in more unfamilier files during the release period. Before the release period main development works are accepted from different branches and they get merged together with main kernel, so, core feature developers finishes their duty for a particular feature development for a particular release within the time of merge period althoug some contributors still work during the release period where basically they work on fixing, testing and minor enhancement related work on the codes that have been merged and accepted from various developers from other branches. It is clear now that during the release period developers work mostly on others' works, the proportion of working with owned files is lower around the release than the proportion during normal development time. When they feel the confidence that everything is tested and can be pushed out for the release, they publish the stable release.

\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3in]{ownpMPRPbox.png}
\caption{\small \sl Percentage of working in owned files during Merge Period and Release Period}
\end{center}
\end{figure}

@@

\subsection{Time for commits to go live}
\textit{How long a commit take to get pushed into the main kernel? Does it differ in different periods of release?}

There are thousands of commits goes to the main kernel during the merge meriod. All they are to be considered as the normal development commits or regular development or core development commits for developing new features to the kernel. Merge period remains opened for two weeks only, the development works that are supposed to go for a particular release are not developed during this couple of weeks. These have been developed since many days ago on a different development branch by any developer. For a particular feature that is going to the current release there are hundreds of commits, obviously all these commits are not made during the merge period of the current release 90-100\% commits must have been made before the opening of the merge period of current release. As described in section 4.3, we have sorted up the releases according to the git DAG. This helps us to calculate how many days in past the first commit of this particular feature had been made which is going to be pushed into the current release. The duration a commit was made to the local branch to the date it is pushed into the main branch is being called here ``lag'' of commits. Figure 10 represents lag of the commits made for regular development and lag of the commits made during the release period.

\begin{figure}
\begin{center}
\includegraphics[height=1.7in,width=3in]{lagMPRPbox.png}
\caption{\small \sl Time takes for development commits and commits around release periods to get released}
\end{center}
\end{figure}

From figure 10 we can understand it easily that rather quick fixes are made during the release period. With the small sets of bug fixes, important wuick enhancements candidate releases are pushed and once the kernel seems to me more stable with all the new features that's been merged at the beginning of the release cycle, stable release is published. Comperatively lag for the feature development commits is quite high. Median for the lags for the commits during this time is 5.57 while median for the lags for the commits goes to the main kernel during merge period is 26.15 which is a significant different between these two periods of a release cycle.
